Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-10 MS_revised.tex
ref 35

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies {[}13{]} to investigate patterns
of diversification. We applied MEDUSA {[}16,17{]}, a recently developed
statistical method, to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA {[}16{]}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} {[}18{]}.

@@

For analyses, we used the phylogenetic trees from the study of Wahlberg
et al. (2009) {[}13{]} that were generated using DNA sequence data from
10 gene regions for 398 of the 540 valid genera in Nymphalidae. We
employed the maximum clade credibility tree (MCC tree) (Fig. 1) as well
as a random sample of 1000 trees from their BEAST run after burnin
{[}13{]}. Their original BEAST run was for 40 million generations. We
used a burnin of 25 million generations and took a random sample of 1000
trees using Burntrees v.0.1.9 \url{http://www.abc.se/~nylander/} (supp.
mat. 5) in order to correct for phylogenetic uncertainty when performing
the diversification analyses.

Species richness data for Nymphalidae genera were compiled from several
sources including the specialist-curated lists on
\url{http://tolweb.org}, Lamas (2004) {[}19{]} and curated lists of
Global Butterfly Names project (\url{http://www.ucl.ac.uk/taxome/gbn/}).
We assigned the species numbers of genera not included in the phylogeny
to the closest related genus that was included in Wahlberg et al. (2009)
{[}13{]} study according to available phylogenetic studies {[}20--31{]},
taxonomical classification and morphological resemblance when no
phylogenies were available (supp. mat. 3).

Hostplant data for Nymphalidae species were compiled from several
sources including {[}32{]}, HOSTS database (\url{http://bit.ly/YI7nwW}),
{[}33--35{]} and others (supp. mat. 15--16) for a total of 6586
hostplant records, including 428 Nymphalidae genera and 143 plant
families and 1070 plant genera. It was not possible to find any
hostplant data for 35 butterfly genera.

>>>>

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies \cite{wahlberg2009} to
investigate patterns
of diversification. We applied MEDUSA \cite{alfaro2009,harmon2011},
a recently developed
statistical method, to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA \cite{alfaro2009}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}.

@@

For analyses, we used the phylogenetic trees from the study of Wahlberg
et al. (2009) \cite{wahlberg2009} that were generated using DNA sequence data from
10 gene regions for 398 of the 540 valid genera in Nymphalidae. We
employed the maximum clade credibility tree (MCC tree) (Fig. 1) as well
as a random sample of 1000 trees from their BEAST run after burnin
\cite{wahlberg2009}. Their original BEAST run was for 40 million generations. We
used a burnin of 25 million generations and took a random sample of 1000
trees using Burntrees v.0.1.9 \url{http://www.abc.se/~nylander/} (supp.
mat. 5) in order to correct for phylogenetic uncertainty when performing
the diversification analyses.

Species richness data for Nymphalidae genera were compiled from several
sources including the specialist-curated lists on
\url{http://tolweb.org}, Lamas (2004) \cite{lamas2004} and curated lists of
Global Butterfly Names project (\url{http://www.ucl.ac.uk/taxome/gbn/}).
We assigned the species numbers of genera not included in the phylogeny
to the closest related genus that was included in Wahlberg et al. (2009)
\cite{wahlberg2009} study according to available phylogenetic studies 
\cite{matos2013,brower2010,kodandaramaiah2010,kodandaramaiah2010a,ortiz2013,desilva2010,freitas2004,pena2006,penz1999,silva2008,pena2011,pena2010},
taxonomical classification and morphological resemblance when no
phylogenies were available (supp. mat. 3).

Hostplant data for Nymphalidae species were compiled from several
sources including \cite{ackery1988}, HOSTS database (\url{http://bit.ly/YI7nwW}),
\cite{dyer2002,beccaloni2008,janzen2009} and others (supp. mat. 15--16)
for a total of 6586
hostplant records, including 428 Nymphalidae genera and 143 plant
families and 1070 plant genera. It was not possible to find any
hostplant data for 35 butterfly genera.
